---
title: "Writing a fetch function"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing a fetch function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This guide is for people who maintain a reader (Zarr, netCDF, HDF5, GDAL,
an object store, a database of chunk references) and want its arrays to
behave as ordinary R arrays. The split of work is simple:

* **altarr** owns indexing, chunk planning, batching, the cache,
  materialization, serialization and type coercion.
* **You** own one function: given a set of chunk coordinates, return those
  chunks' values.

The full rules are in `?altarr_contract`. This guide works through them
with code you can run, then sketches real backends.

```{r lib}
library(altarr)
```

## 1. The smallest fetch function

A fetch function is called as `fetch(chunks)`. `chunks` is an integer
matrix: one row per chunk, one column per dimension, 0-based chunk
coordinates. It returns a list with one vector per row, each holding that
chunk's values in column-major order, clipped at the array edge.

The easiest backend to reason about is an ordinary array in memory. It is
also the reference you will test real backends against.

```{r minimal}
ref <- array(as.double(seq_len(10 * 7)), c(10, 7))
cs <- c(4L, 3L)

## the values of one chunk of 'a', given its 0-based chunk coordinates
chunk_of <- function(a, cc, cs) {
  start <- cc * cs + 1L
  end <- pmin(start + cs - 1L, dim(a))   # clip at the edge
  idx <- Map(seq, start, end)
  as.vector(do.call(`[`, c(list(a), idx, list(drop = FALSE))))
}

fetch_mem <- function(chunks) {
  lapply(seq_len(nrow(chunks)), function(r) chunk_of(ref, chunks[r, ], cs))
}

x <- altarr(dim(ref), cs, fetch_mem)
x[cbind(c(1, 10), c(1, 7))]
identical(x[2:9, 3:6], ref[2:9, 3:6])
```

`chunk_of()` is the only geometry you need. Every other backend in this
guide is the same loop with a different way of getting one chunk's values.

## 2. Test against a reference

altarr's own tests compare every path with base R using `identical()`.
Do the same for your backend: read a small store both ways, once through
your fetch function and once eagerly, and compare across the paths base
R takes.

```{r check}
check_fetch <- function(x, ref) {
  d <- dim(ref)
  mid <- lapply(d, function(n) unique(c(1L, n %/% 2L, n)))  # corners and middle
  pts <- as.matrix(expand.grid(mid))
  c(
    elt        = identical(x[[length(ref)]], ref[[length(ref)]]),
    subset     = identical(x[c(1, length(ref), 3)], ref[c(1, length(ref), 3)]),
    matrix_idx = identical(x[pts], ref[pts]),
    rectangle  = identical(do.call(`[`, c(list(x), mid)),
                           do.call(`[`, c(list(ref), mid))),
    extract    = identical(do.call(altarr_extract, c(list(x), mid)),
                           do.call(`[`, c(list(ref), mid))),
    sum        = identical(sum(x), sum(ref)),
    mean       = identical(mean(x), mean(ref)),
    anyNA      = identical(anyNA(x), anyNA(ref))
  )
}
check_fetch(x, ref)
```

Choose a reference with chunk counts that do not divide the dimensions
evenly, so every test touches edge chunks. Most fetch bugs live there.

`sum()` and `mean()` accumulate chunk by chunk, so on real floating-point
data they can differ from base R in the last bits. Compare them with
`all.equal()` unless your reference values are integers, as here.

## 3. What altarr checks for you

altarr checks what it can check cheaply, and its errors name the problem.
A fetch that returns a padded edge chunk (as Zarr stores them) is stopped
at the first edge chunk:

```{r padded, error = TRUE}
fetch_padded <- function(chunks) {
  lapply(seq_len(nrow(chunks)), function(r) rep(0, prod(cs)))
}
altarr(dim(ref), cs, fetch_padded)[10, 7]
```

So is a list of the wrong length, or a chunk of the wrong kind:

```{r wrong, error = TRUE}
altarr(dim(ref), cs, function(chunks) list(1))[cbind(c(1, 10), c(1, 7))]
altarr(dim(ref), cs, function(chunks) lapply(seq_len(nrow(chunks)), function(r) "a"))[1]
```

An error inside your fetch function surfaces as an ordinary R error from
wherever the read happened, whether that was `[`, `sum()` or `mean()`:

```{r fails, error = TRUE}
sum(altarr(dim(ref), cs, function(chunks) stop("store unreachable")))
```

What altarr cannot check is the values themselves: a chunk in the wrong
order, or the wrong chunk, has the right length. That is what section 2
is for.

## 4. One call, many chunks

altarr hands your fetch function as many chunks per call as it can plan.
A fetch that logs `nrow(chunks)` shows how each base R operation arrives
at your backend:

```{r batches}
make_logged <- function(a, cs) {
  force(a); force(cs)
  calls <- integer(0)
  fetch <- function(chunks) {
    calls[[length(calls) + 1L]] <<- nrow(chunks)
    lapply(seq_len(nrow(chunks)), function(r) chunk_of(a, chunks[r, ], cs))
  }
  list(fetch = fetch,
       take = function() { out <- calls; calls <<- integer(0); out })
}

big <- array(as.double(seq_len(60 * 40 * 12)), c(60, 40, 12))
lg <- make_logged(big, c(10L, 10L, 4L))   # 6 x 4 x 3 = 72 chunks
y <- altarr(dim(big), c(10L, 10L, 4L), lg$fetch)

ops <- list(
  "y[1:5]"                = function() y[1:5],
  "y[cbind(...)] 3 points" = function() y[cbind(c(1, 60, 30), c(1, 40, 20), c(1, 12, 6))],
  "y[1:25, 1:25, 1]"      = function() y[1:25, 1:25, 1],
  "altarr_extract(same)"  = function() altarr_extract(y, 1:25, 1:25, 1),
  "sum(y)"                = function() sum(y),
  "mean(y)"               = function() mean(y)
)
for (nm in names(ops)) {
  altarr_reset(y)   # empty cache, so every operation starts cold
  invisible(lg$take())
  invisible(ops[[nm]]())
  cat(sprintf("%-24s fetch calls: %-3d chunks per call: %s\n", nm,
              length(cl <- lg$take()), paste(cl, collapse = " ")))
}
```

Matrix indexing and `altarr_extract()` arrive as one call. `x[i, j, k]`
arrives one chunk per call, because base R asks for it element by element
(the gap the base R proposal is about). `sum()`, `min()`, `max()` and
materialization arrive in batches of `getOption("altarr.batch_chunks")`
chunks. `mean()`, `prod()` and `anyNA()` scan the array region by region,
and altarr prefetches one layer at a time: all the chunks that share the
last dimension's chunk coordinate (here 6 x 4 = 24).

So the most useful thing a backend can do is read the rows of one call
**concurrently**. With 64 chunks per call against object storage, a serial
loop pays 64 round trips; a concurrent one pays about one.

```{r concurrent, eval = FALSE}
## sketch: read the chunks of one call in parallel
fetch_parallel <- function(chunks) {
  parallel::mclapply(seq_len(nrow(chunks)),
                     function(r) read_one_chunk(chunks[r, ]),
                     mc.cores = 8)
}
```

Concurrency must stay inside fetch, and only R's main thread may call the R
API. Forked workers (`mclapply`), a multi-handle HTTP client (`curl`'s
multi interface) or a C/C++/Rust library that reads on its own threads and
returns plain buffers are all fine.

## 5. Dimension order: reverse, never transpose

Most stores are C-order (row-major) and list dimensions slowest first. R
is column-major and lists dimensions fastest first. A C-order chunk of
shape `(s1, s2, s3)` holds exactly the same bytes as a column-major chunk
of shape `c(s3, s2, s1)`. So:

* give altarr the reversed shape and reversed chunk shape;
* reverse each row of `chunks` before you build the store's chunk key;
* never transpose a chunk.

`altarr_zarr_v2()` does exactly this, and its source is a compact worked
example: `body(altarr_zarr_v2)`. The result has R's usual orientation for
data that came from netCDF or Zarr. `x[lon, lat, time]` corresponds to the
store's `var[time, lat, lon]`, which is the same convention ncdf4 and
RNetCDF use.

## 6. Types, missing values and unpacking

Pass `type = "double"`, `"integer"` or `"logical"` to `altarr()`.
Return chunks of that type to avoid a copy; other numeric or logical
vectors are coerced as `as.integer()` and friends would.

Decoding belongs in fetch. altarr never sees bytes or encodings, so:

* map fill values and `missing_value` to `NA` inside fetch;
* apply `scale_factor` and `add_offset` inside fetch, and declare the
  array `"double"`;
* decompress, unshuffle and byte-swap inside fetch, or hand the bytes to a
  library that does.

64-bit integers have no R type. Return doubles (exact up to 2^53) and say
so in your documentation.

## 7. Recipes: what `saveRDS()` keeps

`saveRDS()` on an altarr array writes the recipe (dim, chunk shape, type
and the fetch function), never the values. The fetch function travels
with its enclosing environment, so build it in a factory that keeps only
what is needed to re-open the source, and `force()` the arguments:

```{r factory}
make_fetch_store <- function(path, dim, chunk) {
  force(path); force(dim); force(chunk)
  function(chunks) {
    lapply(seq_len(nrow(chunks)), function(r) {
      start <- chunks[r, ] * chunk + 1L
      n <- pmin(chunk, dim - start + 1L)
      rep(nchar(path), prod(n))   # stand-in for reading from 'path'
    })
  }
}
z <- altarr(c(400L, 300L), c(64L, 64L),
            make_fetch_store("s3://bucket/array.zarr", c(400L, 300L), c(64L, 64L)))
f <- tempfile(fileext = ".rds")
saveRDS(z, f)
file.size(f)   # bytes, for 120,000 values
```

Two traps:

* A fetch defined at top level refers to the global environment, which
  is saved as a reference, not its contents. The recipe loads in a new
  session, but reading fails with "object not found". Section 1's
  `fetch_mem` is like this: fine in a session, useless saved.
* Connection handles (external pointers: a GDAL dataset, a netCDF id, a
  DBI connection) do not survive `saveRDS()`. Keep a path or URL in the
  recipe and open the handle lazily, re-opening it when it is gone:

```{r handle, eval = FALSE}
make_fetch_handle <- function(path) {
  force(path)
  h <- NULL
  function(chunks) {
    if (is.null(h) || !handle_is_valid(h)) h <<- open_source(path)
    lapply(seq_len(nrow(chunks)), function(r) read_chunk(h, chunks[r, ]))
  }
}
```

## 8. Sources must be pinned

R assumes a vector never changes. altarr re-fetches chunks after cache
eviction, and a recipe re-reads its source when loaded later. If the
source can change underneath, the same element can have two values in one
session. Point fetch at something immutable: an Icechunk snapshot id, a
versioned object, a fixed set of chunk byte references (kerchunk,
blocklist), or content-addressed chunks.

## 9. Sketches for real backends

These are not run here, because altarr does not depend on these packages.
Each is the section 1 loop with a different `chunk_of()`.

**ncdf4.** `ncvar_get()` already uses R's dimension order and applies
scale and offset, so there is no reversal and the array is double:

```{r ncdf4, eval = FALSE}
altarr_ncdf4 <- function(file, varname) {
  force(file); force(varname)
  nc <- ncdf4::nc_open(file)
  v <- nc$var[[varname]]
  d <- v$varsize
  cs <- v$chunksizes
  if (all(is.na(cs))) cs <- d                  # contiguous: one chunk
  cs <- rev(cs)                                # chunksizes are in C order
  ncdf4::nc_close(nc)
  fetch <- function(chunks) {
    nc <- ncdf4::nc_open(file)
    on.exit(ncdf4::nc_close(nc))
    lapply(seq_len(nrow(chunks)), function(r) {
      start <- chunks[r, ] * cs + 1L
      count <- pmin(cs, d - start + 1L)
      as.vector(ncdf4::ncvar_get(nc, varname, start = start, count = count,
                                 collapse_degen = FALSE))
    })
  }
  altarr(d, cs, fetch)
}
```

**gdalraster (classic raster).** GDAL reads a window left to right, top to
bottom. As a column-major array that is `dim = c(ncol, nrow)`, which is
the same reversal, so the GDAL block size is the chunk shape:

```{r gdal, eval = FALSE}
altarr_gdal <- function(dsn, band = 1L) {
  force(dsn); force(band)
  ds <- methods::new(gdalraster::GDALRaster, dsn)
  d <- c(ds$getRasterXSize(), ds$getRasterYSize())
  cs <- ds$getBlockSize(band)
  ds$close()
  h <- NULL
  fetch <- function(chunks) {
    if (is.null(h) || !h$isOpen()) h <<- methods::new(gdalraster::GDALRaster, dsn)
    lapply(seq_len(nrow(chunks)), function(r) {
      off <- chunks[r, ] * cs
      n <- pmin(cs, d - off)
      h$read(band, off[1], off[2], n[1], n[2], n[1], n[2])
    })
  }
  altarr(d, cs, fetch)
}
```

**Chunk byte references** (kerchunk, blocklist, Icechunk manifests). The
recipe holds a reference table; fetch looks up each chunk's
`(url, offset, length)`, issues the range requests concurrently, and
decodes. Reference tables are pinned by construction, which makes them the
best fit for section 8.

## 10. Tools while developing

* `altarr_stats(x)`: counters for every path (element reads, planned
  subsets, fetch calls, chunks fetched, materializations, evictions) and
  the cache state. Use `altarr_reset(x)` between experiments.
* `altarr_plan(x, i, j, k)`: the chunks a rectangular subscript touches,
  as a data frame. It is useful for checking chunk keys by hand.
* `options(altarr.max_materialize = 0)`: refuse all materialization, so
  any operation that would read the whole array errors instead. Run your
  workflow under it to find what materializes.
* `options(altarr.cache_bytes = 0)`: chunks are evicted as soon as each
  request is done, so every read goes to fetch. This is a stress test for
  the backend.
* `options(altarr.batch_chunks = n)`: match batch size to your backend's
  concurrency.

Which base R operations are planned, element by element, or materializing
is measured in `?altarr_contract`.
