---
title: "altarr walkthrough: what base R asks a lazy array for"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{altarr walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

An altarr array is an ordinary R array (a vector with a `dim` attribute
and no class) whose values live in chunks somewhere else. Base R does not
know it is lazy. This walkthrough runs ordinary R code against one and
shows, after each step, what R actually asked the array for.

For the background, and the case for a small change in base R, read the
post [What R does today with an ALTREP that has a dim and no
class](https://hypertidy.org/posts/2026-10-01_altrep-dim-no-class/). The
full contract for fetch functions is in `?altarr_contract`.

## The example store

`altarr_example_zarr()` writes a small Zarr v2 store (70 x 50 x 30,
chunks 20 x 16 x 7, zlib compressed) to a temporary directory. Every value
is its own 0-based position, so any result can be checked by eye:
`x[i, j, k] == (i-1) + 70*(j-1) + 3500*(k-1)`.

After each step, `seen()` prints the counters that moved since the
previous step:

* `elt`: single-element reads (the `x[i, j, k]` path)
* `extract_subset`: planned subset requests (`x[i]`, `x[cbind(...)]`)
* `hyperslab`: `altarr_extract()` calls
* `fetch_calls`: calls to the fetch function, that is, round trips to the store
* `chunks_fetched`: chunks those calls returned
* `materialize`: whole-array reads (printing, arithmetic, ...)

```{r seen}
library(altarr)

seen <- function(x) {
  s <- altarr_stats(x)
  keep <- c("elt", "extract_subset", "hyperslab", "fetch_calls",
            "chunks_fetched", "materialize")
  print(s[keep])
  cat(sprintf("chunks now cached: %d of %d   materialized: %s\n",
              s[["chunks_cached"]], s[["chunks_total"]],
              if (s[["materialized"]] == 1) "YES" else "no"))
  altarr_reset(x, cache = FALSE)   # zero the counters, keep the cache
  invisible(s)
}
```

## 1. Opening a store reads nothing

```{r open}
path <- altarr_example_zarr()
x <- altarr_zarr_v2(path)
dim(x)
length(x)
class(x)    # "array": no class of its own
seen(x)
```

## 2. One value reads one chunk

```{r one}
x[1]
seen(x)
```

## 3. The cache: reading it again is free

```{r cache}
x[1:5]
seen(x)
```

## 4. Points: `x[cbind(i, j, k)]` is one planned request

Matrix indexing reaches altarr as a single request, so all the chunks it
touches go to the store in one fetch call.

```{r points}
pts <- cbind(c(70, 35, 1), c(50, 25, 1), c(30, 15, 30))
x[pts]
seen(x)
```

## 5. A rectangle: `x[i, j, k]` reads element by element

Base R's `ArraySubset` asks for each element in turn. The result is
correct and still lazy (each chunk is fetched once, then served from the
cache), but there is one round trip per chunk and no chance to plan.

```{r rect}
altarr_reset(x)   # drop the cache too, for a fair count
r1 <- x[1:40, 1:20, 1:3]
seen(x)
```

## 6. The same rectangle, planned

`altarr_extract()` takes the same subscripts and plans the read. This is
what a hook in R's `ArraySubset` would give `x[i, j, k]` for free.

```{r extract}
altarr_reset(x)
r2 <- altarr_extract(x, 1:40, 1:20, 1:3)
identical(r1, r2)
seen(x)
altarr_plan(x, 1:40, 1:20, 1:3)
```

## 7. Materialization: when R wants all the data

Some operations need a pointer to the whole array: printing, arithmetic,
most of `apply()`. Then altarr reads every chunk and keeps the full array.
The operation triggers this, not the size. The option
`altarr.max_materialize` (default 1e6 values) only decides whether it is
allowed. This array has 105,000 values, under the limit.

```{r materialize}
altarr_reset(x)
y2 <- x * 2
seen(x)   # batches of getOption("altarr.batch_chunks", 64)
x[1:3]    # from memory now: no counters move
seen(x)
```

Typing `x` at the console would do the same, because `print()` asks for
the whole array.

## 8. With a lower limit, materializing is refused

```{r refuse}
x2 <- altarr_zarr_v2(path)
op <- options(altarr.max_materialize = 1e4)
try(x2 * 2)
x2[1:3]   # indexing still works
seen(x2)
options(op)
```

## 9. Copies stay lazy until changed

```{r copies}
x3 <- altarr_zarr_v2(path)
y <- x3                  # no copy, no read
y[1, 1, 1] <- -1         # changing y materializes y only
c(y[1, 1, 1], x3[1, 1, 1])
seen(x3)
```

A copy shares its original's recipe, counters and chunk cache, so the
counters include `y`'s materialization. Materializing fills `y` directly,
not through the shared cache, so `x3`'s cache holds only the chunk `x3`
itself read. The cache is bounded: see `getOption("altarr.cache_bytes")`
and `evictions` in `altarr_stats()`.

## 10. Saving stores the recipe, not the data

```{r save}
f <- tempfile(fileext = ".rds")
saveRDS(x3, f)
file.size(f)   # the path, shape and chunking, not 105,000 values
x4 <- readRDS(f)
x4[pts]
```

## Where to go next

* `?altarr_contract`: what a fetch function must do, which base R
  operations are planned, element-wise or materializing, and the options.
* `?altarr`: build a lazy array over any store from a fetch function.
* The [hypertidy
  post](https://hypertidy.org/posts/2026-10-01_altrep-dim-no-class/): the
  proposal for base R.
